GEE models

### 2016-07-27 doc/manuscript.Rmd
Added tony's edits

Table and figure count: 3 / 4

@@

Participants from London and Toronto, Canada, were recruited into the 
Prospective Metabolism and Islet cell Evaluation (PROMISE) cohort. Eligibility 
for recruitment into PROMISE required having one or more risk factors for type 2
diabetes mellitus, including obesity, hypertension, family history of diabetes, 
and/or a history of gestational diabetes or birth of a macrosomic infant. 
Participants aged 30 years and older (n=736) attended the baseline visit
between 2004-2006. Follow-up examinations in this cohort occur every three years, with three
examination visits completed to date (2004-2006, 2007-2009, and 2010-2013). 
Participants are contacted annually by telephone. The
current study used data on participants who did not have diabetes at baseline,
who returned for one or more of the follow-up examinations, and who had samples
available for fatty acid measurements (n=477). A diagram of the sample size at
each visit is shown in `r cite_sf('consort')`. At each examination,
participants undergo metabolic characterization, anthropometric measurements,
and questionnaires on lifestyle and sociodemographics. Research ethics approval
was obtained from Mount Sinai Hospital and the University of Western Ontario,
and all participants provided written informed consent.
Data collection methods were standardized across the 2 centres and research
nurses were centrally trained.

@@

In a Canadian population of adults who are at-risk for diabetes, we found that 
higher total NEFA concentrations independently predicted lower beta-cell
function after 6 years. While we found negative associations with palmitic acid
(16:0), oleic acid (18:1n-9), and linoleic acid (18:2n-6) for the concentration
(nmol/mL) modeling, no associations were seen for these fatty
acids as a proportion, an observation that was further confirmed by multivariate
cluster techniques, which found no predictive ability of these
individual NEFA on beta-cell function or dysglycemia. These observations suggest
that the absolute size of the total NEFA fraction, rather than its specific
composition, likely influences the pathogenesis of diabetes, at least within a
population at-risk for diabetes.

There are a few important limitations to our study. NEFA were only quantified at
the baseline visit and as such we cannot investigate whether there were 
concomitant changes in NEFA and the metabolic measures. However, we believe this
is a strength for our specific objective, as the chance of reverse causality is 
reduced given that fatty acid and glucose metabolism pathways are tightly 
integrated. This is also an observational cohort, and there may be some residual
confounding we have not considered or that could not be measured. Nonetheless,
potential covariates were empirically analyzed prior to inclusion into the
GEE models to best understand and minimize potential
confounding. Finally, our cohort consists of individuals at-risk for diabetes, 
who are primarily female and European-ancestry and as such our results may not be
generalizable to other populations. However, given these limitations, our study
also has several strengths, including the longitudinal design and the rigorous
statistical techniques and methods applied, which are specifically suited to
investigating temporal relationships and to handling the multivariate nature of
the data. Lastly, our cohort contains highly detailed and comprehensive
variable measurements at each collection visit, and has both concentration and
mol% data for the fatty acids.

@@

In conclusion, we found that total NEFA was a strong predictor for lower
beta-cell function over 6 years, irrespective of the specific composition of the
NEFA fraction, suggesting that efforts at lower circulating NEFA
indiscriminately to reduce the risk of diabetes are well warranted. While future
studies are needed to confirm these findings, our results reinforce the
importance of continuing to investigate the role of circulating NEFA
concentration on the natural history of diabetes.

>>>>

<!-- Table and figure count: 3 / 4 -->

@@

Participants from London and Toronto, Canada, were recruited into the 
Prospective Metabolism and Islet cell Evaluation (PROMISE) cohort. Eligibility 
for recruitment into PROMISE required having one or more risk factors for type 2
diabetes mellitus, including obesity, hypertension, family history of diabetes, 
and/or a history of gestational diabetes or birth of a macrosomic infant. 
Participants aged 30 years and older (n=736) attended the baseline visit
between 2004-2006. Follow-up examinations in this cohort occur every three years, with three
examination visits completed to date (2004-2006, 2007-2009, and 2010-2013). 
Participants are contacted annually by telephone. The
current study sample used data on participants who did not have diabetes at baseline,
who returned for one or more of the follow-up examinations, and who had samples
available for fatty acid measurements (n=477). A diagram of the sample size at
each visit is shown in `r cite_sf('consort')`. At each examination,
participants undergo metabolic characterization, anthropometric measurements,
and questionnaires on lifestyle and sociodemographics. Research ethics approval
was obtained from Mount Sinai Hospital and the University of Western Ontario,
and all participants provided written informed consent.
Data collection methods were standardized across the 2 centres and research
nurses were centrally trained.

@@

In a Canadian population of adults who were at-risk for diabetes, we found that 
higher total NEFA concentrations independently predicted lower beta-cell
function after 6 years. While we found negative associations with palmitic acid
(16:0), oleic acid (18:1n-9), and linoleic acid (18:2n-6) for the concentration
(nmol/mL) modeling, no associations were seen for these fatty
acids as a proportion, an observation that was further confirmed by multivariate
cluster techniques, which found no ability of these
individual NEFA to predict beta-cell function or dysglycemia. These observations suggest
that the absolute size of the total NEFA fraction, rather than its specific
composition, likely influences the pathogenesis of diabetes, at least within a
population at-risk for diabetes.

There are a few important limitations to our study. NEFA were only quantified at
the baseline visit and as such we cannot investigate whether there were 
concomitant changes in NEFA and the metabolic measures over time. However, we believe this
is a strength for our specific objective, as the chance of reverse causality is 
reduced given that fatty acid and glucose metabolism pathways are tightly 
integrated. This is also an observational cohort, and there may be some residual
confounding we have not considered or that could not be measured. Nonetheless,
potential covariates were empirically analyzed prior to inclusion into the
GEE models to best understand and minimize potential
confounding. Finally, our cohort consists of individuals at-risk for diabetes, 
who are primarily female and European-ancestry and as such our results may not be
generalizable to other populations. However, given these limitations, our study
also has several strengths, including the longitudinal design and the rigorous
statistical techniques and methods applied, which are specifically suited to
investigating temporal relationships and to handling the multivariate nature of
the data. Lastly, our cohort contains highly detailed and comprehensive
variable measurements at each collection visit, and has both concentration and
mol% data for the fatty acids.

@@

In conclusion, we found that total NEFA was a strong predictor for lower
beta-cell function over 6 years, irrespective of the specific composition of the
NEFA fraction, suggesting that efforts at lowering circulating total NEFA
through medication and/or lifestyle strategies to reduce the risk of diabetes
are well warranted. While future
studies are needed to confirm these findings, our results reinforce the
importance of continuing to investigate the role of circulating NEFA
concentration on the natural history of diabetes.
